The Calculus of Linear Constructions

### 2021-10-11 writing/draft1.tex
wip

\renewcommand{\syntleft}{}
\renewcommand{\syntright}{}
\setlength{\grammarparsep}{10pt}
\setlength{\grammarindent}{10em}
\newcommand{\indalt}[1][2]{\\\hspace*{-1.2em}\textbar\quad}
\newcommand{\rname}[1]{\textsc{\footnotesize #1}}
\newcommand{\LNLD}{\text{LNL}_\text{D}}
\newcommand{\ok}{\textbf{ok}}
\newcommand{\pure}[1]{|#1|}
\newcommand{\sort}{\textbf{sort}}
\newcommand{\type}{\textbf{type}}
\newcommand{\refl}{\text{refl}}
\newcommand{\letin}[3]{$\text{let }#1\text{ := }#2\text{ in }#3$}
\newcommand{\new}{\text{new}}
\newcommand{\free}{\text{free}}
\newcommand{\get}{\text{get}}
\newcommand{\set}{\text{set}}
\newcommand{\subst}{\text{subst}}
\newcommand{\utype}{:_U}
\newcommand{\ltype}{:_L}
\newcommand{\stype}[1]{:_#1}
\newcommand{\step}{\leadsto}
\newcommand{\pstep}{\leadsto}
\newcommand{\mrg}[3]{#1\ddagger#2\ddagger#3}
\newcommand{\arw}[2]
{\hspace*{0.2em}{\scriptstyle #1}\hspace*{-0.2em}
\rightarrow
\hspace*{-0.2em}{\scriptstyle #2}\hspace*{0.2em}}

@@

      <$m, n, A, B, C$> ::= $U_i$ | $L_k$ | $x$ \hspace*{7.3em} expressions
      \indalt $(x : A) \arw{s}{t} B$
      \indalt $\lambda x. n$ | $m\ n$
    \end{grammar}
    \end{minipage}
    \label{syntax}
  \end{figure}
  A clear departure of our language from standard presentations of both linear type theory, and dependent type theory is the indexed function type: $(x : A) \arw{s}{t} B$. The reason for these indices is that we have built the ! exponential of linear logic directly into universe sorts. The indices $s$, and $t$ annotate the domain, and co-domain's respective universe sort, hence annotating linearity. The behavior of ! is difficult to account for even in simple linear type theory. Subtle issues arise if !! is not canonically isomorphic to !, which may invalidate the substitution lemma \cite{substitute}. By integrating the exponential directly into type formation rules, we implicitly limit ! to only be used canonically. This allows us to derive the substitution lemma, and construct a direct modeling of CLC in CC$\omega$ without needing extra machinery for manipulating exponential.
  \begin{figure}[H]
    \caption{Correspondence of CLC types and MELL implications}
    \begin{align}
      (\_ : A) \arw{U}{U} B \quad &\equiv \quad !A \multimap !B \\
      (\_ : A) \arw{U}{L} B \quad &\equiv \quad !A \multimap B \\
      (\_ : A) \arw{L}{U} B \quad &\equiv \quad A \multimap !B \\
      (\_ : A) \arw{L}{L} B \quad &\equiv \quad A \multimap B
    \end{align}
    \label{correspondence}
  \end{figure}
  Figure \ref{correspondence} illustrates the correspondence between CLC function types and Multiplicative Exponential Linear Logic (MELL) implications. MELL lacks counterparts for the cases (1), (2) if the co-domain $B$ is dependent on arguments of the domain $A$. We will discuss the detail of these function types in Section \ref{tyformation}.

@@

  \subsection{Typing Judgment}
  Typing judgments in CLC take on the form of $\Gamma \vdash m \stype{s} A$. Intuitively, this judgment states that the term $m$ is an inhabitant of type $A$, with free variables typed in $\Gamma$. The sort index $s$ tells us the linearity of $m$. Specifically, if $s = U$, then $m$ has unrestricted usage. Conversely, if $s = L$, then $m$ must be used exactly once. In Section \ref{meta}, we show through the regularity theorem that $s$ corresponds exactly to the sort of type $A$'s universe.

  \subsection{Type Formation} \label{tyformation}
  The rules for forming types are presented in Figure \ref{type}. In CLC, we forbid types from depending on linear terms similar to \cite{llf,neel15} for the same reason of avoiding philosophical pitfalls.

@@

  Recall from Figure \ref{correspondence} where we presented the correspondence between CLC function types and MELL implications. There exists 8 canonical combinations of ! and $\multimap$ that we must encode. Note that the function types $(x : A) \rightarrow_s B$, and $A \rightarrow_s B$ are intrinsically different entities, even if the variable $x$ is unbinding in the former. So $A \rightarrow_s B$ should not be viewed as a shorthand for $(\_ : A) \rightarrow_s B$. Likewise for the linear variants, $A \multimap_s B$ is not a shorthand for $(\_ : A) \multimap_s B$.

  The \rname{U-Prod} rule is used for forming non-linear dependent function types. $\lambda$-abstractions of this type could be applied an unrestricted number of times, with the argument being allowed unrestricted usage within the function body as well. Notice that the domain $A$ must be of type $U_k$, this entails that the co-domain $B$ may depend on an unrestricted term of type $A$. However, $B$ itself may chose to be of any universe sort with an appropriate predicative level. The sort of $B$'s universe is used to annotate the function arrow. The side-condition $\pure{\Gamma}$ ensures that all components used to form $(x : A)\rightarrow_s B$ do not dependent on linear variables, thus allowing the \textit{type} itself to used as an unrestricted \textit{term}.

  The \rname{Arrow} rule is also used for forming non-linear function types. However, as we have alluded to previously, the formation rule \rname{Arrow} differs from \rname{U-Prod}. Namely, the domain $A$ must be of type $L_k$. This means that if the co-domain $B$ is dependent on a term of type $A$, $B$ will depend on a linear term. To prevent this situation from ever occurring, we explicitly forbid the binding of a dependent variable $x$ within $B$. This leaves us with a type of the shape $A \rightarrow_s B$, where $s$ is the sort of $B$'s universe. $\lambda$-abstractions of this type could be applied an unrestricted number of times, but the argument must be used within the body of the function once.

  The \rname{L-Prod} rule is used for forming linear function types. $\lambda$-abstractions of this type must be applied exactly once, with the argument being allowed unrestricted usage within the function. \rname{L-Prod} shares the same

>>>>

\renewcommand{\syntleft}{}
\renewcommand{\syntright}{}
\setlength{\grammarparsep}{10pt}
\setlength{\grammarindent}{10em}
\newcommand{\indalt}[1][2]{\\\hspace*{-1.2em}\textbar\quad}
\newcommand{\rname}[1]{\textsc{\footnotesize #1}}
\newcommand{\LNLD}{\text{LNL}_\text{D}}
\newcommand{\ok}{\textbf{ok}}
\newcommand{\pure}[1]{|#1|}
\newcommand{\sort}{\textbf{sort}}
\newcommand{\type}{\textbf{type}}
\newcommand{\refl}{\text{refl}}
\newcommand{\letin}[3]{$\text{let }#1\text{ := }#2\text{ in }#3$}
\newcommand{\new}{\text{new}}
\newcommand{\free}{\text{free}}
\newcommand{\get}{\text{get}}
\newcommand{\set}{\text{set}}
\newcommand{\subst}{\text{subst}}
\newcommand{\utype}{:_{\scriptscriptstyle U}}
\newcommand{\ltype}{:_{\scriptscriptstyle L}}
\newcommand{\stype}[1]{:_#1}
\newcommand{\step}{\leadsto}
\newcommand{\pstep}{\leadsto}
\newcommand{\mrg}[3]{#1\ddagger#2\ddagger#3}
\newcommand{\arw}[2]
{\hspace*{0.2em}{\scriptstyle #1}\hspace*{-0.2em}
\rightarrow
\hspace*{-0.2em}{\scriptstyle #2}\hspace*{0.2em}}

@@

      <$m, n, A, B, C$> ::= $U_i$ | $L_k$ | $x$ \hspace*{7.3em} expressions
      \indalt $(x : A) \arw{s}{t} B$
      \indalt $\lambda x. n$ | $m\ n$
    \end{grammar}
    \end{minipage}
    \label{syntax}
  \end{figure}
  A clear departure of our language from standard presentations of both linear type theory, and dependent type theory is the indexed function type: $(x : A) \arw{s}{t} B$. The reason for these indices is that we have built the ! exponential of linear logic directly into universe sorts. The indices $s$, and $t$ annotate the domain, and co-domain's respective universe sort, hence annotating linearity. The behavior of ! is difficult to account for even in simple linear type theory. Subtle issues arise if !! is not canonically isomorphic to !, which may invalidate the substitution lemma \cite{substitute}. By integrating the exponential directly into type formation rules, we implicitly limit ! to only be used canonically. This allows us to derive the substitution lemma, and construct a direct modeling of CLC in CC$\omega$ without needing extra machinery for manipulating exponential.
  \begin{figure}[H]
    \caption{Correspondence of CLC types and MELL implications}
    \begin{align}
      (\_ : A) \arw{U}{U} B \quad &\equiv \quad !A \multimap !B \\
      (\_ : A) \arw{U}{L} B \quad &\equiv \quad !A \multimap B \\
      (\_ : A) \arw{L}{U} B \quad &\equiv \quad A \multimap !B \\
      (\_ : A) \arw{L}{L} B \quad &\equiv \quad A \multimap B
    \end{align}
    \label{correspondence}
  \end{figure}
  Figure \ref{correspondence} illustrates the correspondence between CLC function types and Multiplicative Exponential Linear Logic (MELL) implications. MELL lacks counterparts for the cases (1), (2) if the co-domain $B$ is dependent on arguments of the domain $A$. Though it may seem as if implications of the $!(A \multimap B)$ form are left out, CLC encodes these by endowing each function type with a universe sort in their formation judgment. The encoding for this example would be $\Gamma \vdash A \arw{L}{L} B \utype U_i $. We will discuss function type formation in Section \ref{tyformation} in more detail.

@@

  \subsection{Typing Judgment}
  Typing judgments in CLC take on the form of $\Gamma \vdash m \stype{s} A$. Intuitively, this judgment states that the term $m$ is an inhabitant of type $A$, with free variables typed in $\Gamma$. The sort index $s$ tells us the linearity of $m$. Specifically, if $s = U$, then $m$ has unrestricted usage. Conversely, if $s = L$, then $m$ must be used exactly once. In Section \ref{meta}, we show through the regularity theorem that $s$ corresponds exactly to the sort of $A$'s universe.

  \begin{definition} We formally define the terms \textit{non-linear}, \textit{linear}, \textit{unrestricted}, and \textit{restricted}.
    \begin{enumerate}
      \item A \textit{type} $A$ is \textit{non-linear} under context $\Gamma$ if $\Gamma \vdash A \utype U_i$.
      \item A \textit{type} $A$ is \textit{linear} under context $\Gamma$ if $\Gamma \vdash A \utype L_k$.
      \item A \textit{term} $m$ is \textit{unrestricted} under context $\Gamma$ if it has type $A$, and $\Gamma \vdash m \utype A$.
      \item A \textit{term} $m$ is \textit{restricted} under context $\Gamma$ if it has type $A$, and $\Gamma \vdash m \ltype A$
    \end{enumerate}
  \end{definition}

  \subsection{Type Formation} \label{tyformation}
  The rules for forming types are presented in Figure \ref{type}. In CLC, we forbid types from depending on linear terms similar to \cite{llf,neel15} for the same reason of avoiding philosophical troubles.

@@

  The \rname{U-Prop} rule is used for forming propositions. From the judgment $\Gamma \vdash A \utype U_i$ we can see that \rname{U-Prop} allows for quantification over non-linear types of arbitrary level, hence impredicative. The judgment $\Gamma, x \utype A \vdash B \utype U_*$ asserts that the co-domain must be in the impredicative universe $U_*$. The final resulting judgment $\Gamma \vdash (x : A) \arw{U}{U} B \utype U_*$ is indexed by $U$, indicating that the \textit{type} $(x : A) \arw{U}{U} B$ can be use unrestrictedly as a \textit{term}. Since $(x : A) \arw{U}{U} B$ belongs to universe $U_*$, its \textit{terms} also enjoys unrestricted usage.

  The \rname{U-Prod} rule is used for forming function types with non-linear domains. This is evident from the judgment $\Gamma \vdash A \utype U_k$. Since $A$ is in the predicative non-linear universe $U_k$, terms of type $A$ have unrestricted usage. The non-linearity of domain $A$ allows $B$ to depend on terms of type $A$, as seen in judgment $\Gamma, x \utype A \vdash B \utype s_k$. The domain $B$ itself may be non-linear or linear, since $B$'s universe $s_k$ can vary between $U_k$ and $L_k$. From the resulting judgment $\Gamma \vdash (x : A) \arw{U}{s} B \utype t_k$, we see that $A$, and $B$'s universe sorts are used as indices to the arrow. Interestingly, $(x : A) \arw{U}{s} B$ is assigned to a universe $t_k$ of level $k$ and arbitrary sort $t$. If $t$ is chosen to be $U$, then $\Gamma \vdash (x : A) \arw{U}{s} B \utype U_k$ tells us $\lambda$-abstractions of this type have unrestricted usage. Conversely, if $t$ is chose to be $L$, then $\Gamma \vdash (x : A) \arw{U}{s} B \utype L_k$ tells us $\lambda$-abstractions of this type have restricted usage. This is how we encode the missing $!(A \multimap B)$ forms from MELL shown in Figure \ref{correspondence}.

  The \rname{L-Prod} rule is used for forming function types with linear domains. From the judgment $\Gamma \vdash A \utype L_k$,
